Icicle: write once, run once

### 2016-05-18 2016fhpc/section/01-Introduction.tex
Typechecking etc

In Icicle, the above queries are expressed as:
\begin{code}
query stocks
in filter open > close
in count.

query stocks
in filter open < close
in count.

query stocks
in filter open > close
in sum open / count.
\end{code}

>>>>

In Icicle, the above queries are expressed as:
\begin{code}
table stocks { open : Int; close : Int; }
query more = filter open > close in count;
query less = filter open < close in count;
query mean = filter open > close in sum open / count;
\end{code}

### 2016-05-18 2016fhpc/section/02-Source.tex
Typechecking etc

\begin{figure}

\begin{code}
Exp   := x | Exp Op Exp | x Exp... | Query

Query := fold x = Exp then Exp
       | let  x = Exp          in Query
       | filter   Exp          in Query
       | group    Exp          in Query
       | Exp

Top   := query x in Query
Fun   := function x x... = Query
\end{code}

\caption{Icicle Grammar}
\label{fig:icicle:grammar}
\end{figure}

The grammar for Icicle is given in figure~\ref{fig:icicle:grammar}.
@Exp@ and @Query@ are mutually recursive, defining the body of a query.
@Top@ defines a top-level query over a particular table, whereas @Fun@ is used for function definitions.

Expressions can be variable names, operators such as addition and division, function application, or nested queries.
Function application @x Exp...@ is the name of a top-level function applied to any number of arguments.

@@

Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
It returns a fold that starts at zero and for every element, increases the old sum by @value@.
\begin{code}
function sum value
 = fold s = 0 then s + value
\end{code}

The definition of mean is just the sum divided by the count.
\begin{code}
function mean value
 = sum value / count
\end{code}

@@

@@

\begin{code}
   let k    = newest key
in let avgs = group  key in mean value
in avgs ! k
\end{code}

\subsection{Bounded buffer restriction}
As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.
Any operation which requires a buffer must be bounded in size, as an unbounded buffer could potentially grow too large to fit in memory.

>>>>

\input{figures/Source-Grammar.tex}
\input{figures/Source-Type.tex}
\input{figures/Source-Eval.tex}

The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
Value types are given in $T$ and can be numbers, booleans or maps.
Modal types, $T_M$ can be a pure value type, or a modality associated with a value type.
Function types, $T_\to$, can be either a non-function modal type, or a function from modal type to function type.
As Icicle is a first-order language, function types are not value types.

The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
$\mi{Exp}$ defines the expressions used as query and function bodies.
$\mi{Prim}$ defines the primitive types.
Function definitions are given in $\mi{Fun}$.
Query definitions are given in $\mi{Query}$.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.

\TODO{update this to fit new grammar}
Expressions can be variable names, primitive operators such as addition and division, function application, or nested queries.
Function application @x Exp*@ is the name of a top-level function applied to any number of arguments.
Primitive application @Prim Exp*@ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

@@

Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
It returns a fold that starts at zero and for every element, increases the old sum by @value@.
\begin{code}
function sum (value : Element Int)
 = fold s = 0 then s + value
\end{code}

The definition of mean is just the sum divided by the count.
\begin{code}
function mean (value : Element Int)
 = sum value / count
\end{code}

@@

Folds have a name and two expressions.
The first expression, the initialiser, is pure.
The second expression, the update, has the fold's name bound to an @Element@ of the previous value.
The type of the expression must be @Element@.
The final return of the fold is an @Aggregate@.
\begin{code}
fold x = T then Element T : Aggregate T
\end{code}

Let-bindings are independent of the modality.

For filters, the predicate is an @Element@, which could be thought of as a stream of booleans.
The rest of the query must be an @Aggregate@; that is, it must return a scalar rather than a stream, to enforce the bounded buffer restriction explained in \sref{s:IcicleSource:bounded}.
\begin{code}
filter (Element Bool) in (Aggregate T)
\end{code}

Grouping is similar to filters; the key must be an @Element@, while the group value must be an @Aggregate@.
\begin{code}
group (Element Key) in (Aggregate Value)
\end{code}

@@

\begin{code}
   let k    = newest key
in let avgs = group  key in mean value
in mapLookup avgs k
\end{code}

\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.
Any operation which requires a buffer must be bounded in size, as an unbounded buffer could potentially grow too large to fit in memory.

### 2016-05-18 2016fhpc/section/03-Core.tex
Typechecking etc

\begin{code}
Exp   := x | Prim Exp...

Plan  := query  x with (x : T)...
       before { x : T = Exp... }
       folds  { x : T = Exp then Exp... }
       after  { x : T = Exp... }
       return { x : T = x... }

T     := bool | int | map T T
\end{code}

@@

One of the simplest queries to convert to a query plan is a count.
The query plan header specifies the table and the type of each column in the table.
This defines a single fold called @count@ which starts at zero, and is incremented for each element.
At the end of the query, the final count is assigned to the output @count_query@, which is the query's name.
\begin{code}
count_query =
query stocks
in    count

@@

query stocks with (open : Int) (close : Int)
before { }
folds  { count : Int = 0 then count + 1 }
after  { }
return { count_query : Int = count }
\end{code}

Computing the sum is only slightly more complicated.
Here there is again a single fold, @sum@, starting at zero and being incremented by the current column's @open@.
\begin{code}
sum_open =
query stocks
in    sum open

@@

query stocks with (open : Int) (close : Int)
before { }
folds  { sum : Int = 0 then sum + open }
after  { }
return { sum_open : Int = sum }
\end{code}

In order to compute the mean, we compute the sum and the count, and after all computation has finished, divide the two.
Here we use two folds, one for the sum, and one for the count.
While we could have used a single fold containing a pair of the two, keeping the folds separate exposes more opportunities for common subexpression elimination.
\begin{code}
mean_open =
query stocks
in    sum open / count

@@

query stocks with (open : Int) (close : Int)
before { }
folds  { sum       : Int = 0 then sum + open 
         count     : Int = 0 then count + 1 }
after  { mean      : Int = sum / count }
return { mean_open : Int = mean }
\end{code}

If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
The names of the returns and the column names stay the same.
Query plans can only be fused if they act on the same table.
\begin{code}
query stocks with (open : Int) (close : Int)
before { }
folds  { c1 : Int = 0 then c1 + 1
         c2 : Int = 0 then c2 + 1
         s1 : Int = 0 then s1 + open 
         s2 : Int = 0 then s2 + open }
after  { m1 : Int = s2 / c2 }
return { count_query : Int = c1
         sum_open    : Int = s1
         mean_open   : Int = m1 }
\end{code}

The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
We can perform fairly standard common subexpression elimination to remove these.
\begin{code}
query stocks with (open : Int) (close : Int)
before { }
folds  { c1 : Int = 0 then c1 + 1
         s1 : Int = 0 then s1 + open }
after  { m1 : Int = s1 / c1 }
return { count_query : Int = c1
         sum_open    : Int = s1
         mean_open   : Int = m1 }
\end{code}

>>>>

\begin{code}
Exp   := x | Prim Exp*

Plan  := plan   x { (x : T;)* }
       before { x : T = Exp;* }
       folds  { x : T = Exp then Exp;* }
       after  { x : T = Exp;* }
       return { x : T = x;* }
\end{code}

@@

One of the simplest queries to convert to a query plan is a count.
The query plan header specifies the table and the type of each column in the table.
This defines a single fold called @count@ which starts at zero, and is incremented for each element.
At the end of the query, the final count is assigned to the output @count_query@, which is the query's name.
\begin{code}
table stocks { open : Int; close : Int; }
query count_query = count

@@

plan stocks { open : Int; close : Int; }
before { }
folds  { count : Int = 0 then count + 1 }
after  { }
return { count_query : Int = count }
\end{code}

Computing the sum is only slightly more complicated.
Here there is again a single fold, @sum@, starting at zero and being incremented by the current column's @open@.
\begin{code}
query sum_open = sum open

@@

plan stocks { open : Int; close : Int; }
before { }
folds  { sum : Int = 0 then sum + open }
after  { }
return { sum_open : Int = sum }
\end{code}

In order to compute the mean, we compute the sum and the count, and after all computation has finished, divide the two.
Here we use two folds, one for the sum, and one for the count.
While we could have used a single fold containing a pair of the two, keeping the folds separate exposes more opportunities for common subexpression elimination.
\begin{code}
query mean_open = sum open / count

@@

plan stocks { open : Int; close : Int; }
before { }
folds  { sum       : Int = 0 then sum + open 
         count     : Int = 0 then count + 1 }
after  { mean      : Int = sum / count }
return { mean_open : Int = mean }
\end{code}

If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
The names of the returns and the column names stay the same.
Query plans can only be fused if they act on the same table.
\begin{code}
plan stocks { open : Int; close : Int; }
before { }
folds  { c1 : Int = 0 then c1 + 1
         c2 : Int = 0 then c2 + 1
         s1 : Int = 0 then s1 + open 
         s2 : Int = 0 then s2 + open }
after  { m1 : Int = s2 / c2 }
return { count_query : Int = c1
         sum_open    : Int = s1
         mean_open   : Int = m1 }
\end{code}

The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
We can perform fairly standard common subexpression elimination to remove these.
\begin{code}
plan stocks { open : Int; close : Int; }
before { }
folds  { c1 : Int = 0 then c1 + 1
         s1 : Int = 0 then s1 + open }
after  { m1 : Int = s1 / c1 }
return { count_query : Int = c1
         sum_open    : Int = s1
         mean_open   : Int = m1 }
\end{code}
